---
title: "HW 12: Section 4"
author: "Carly Chang (A16843962)"
format: pdf
---

## 4. Population Scale Analysis

One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. So, you processed about
~230 samples and did the normalization on a genome level. Now, you want to find whether
there is any association of the 4 asthma-associated SNPs (rs8067378...) on ORMDL3
expression.

> Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes. Hint: The read.table(),
summary() and boxplot() functions will likely be useful here. There is an example R script
online to be used ONLY if you are struggling in vein. Note that you can find the medium value
from saving the output of the boxplot() function to an R object and examining this object.
There is also the medium() and summary() function that you can use to check your
understanding.


```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt") # get file
head(expr)
```
```{r}
nrow(expr) # number of rows (samples)
```

```{r}
table(expr$geno) # number of each genotype
```



> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3? Hint: An example boxplot is provided overleaf – yours does not need
to be as polished as this one.


```{r}
library(ggplot2)
ggplot(expr) + 
  aes(geno, exp, fill=geno) +
  geom_boxplot(notch=T)
```

If notches don't overlap, then the difference is usually significant. 
G/G and A/A expression levels are quite different, indicating that the SNP affects the expression of ORMDL3.


